---
title: "NIPMAP on HIFI data"
output:
  html_document: default
  pdf_document: default
date: "2023-03-28"
---
```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())

library(tidyverse)
library(RColorBrewer)
library(rjson)
library(readxl)
library(plotly)
library(grid)
library(gridExtra)
dirName <- dirname(rstudioapi::getSourceEditorContext()$path) # get the path
setwd(dirName)

### required functions (NIPMAP tool)
source("../../NIPMAP(Requierd_Packages)/functions_phenotypes_tmens.r")



```

### read the csv files for each group seperatly and save it as an object (untreated, 7DaysRT and Rebound) and combine all groups together as a CSV file.

```{r , Preprocessing ,echo = TRUE}


sample_Un = "31801_32Un" # untreated
sample_Tr = "31803_13Tr" # 7days IR
sample_Re = "33876_36Re" # relapse
Un = "Untreated_31801_slide_32"
Tr = "7DayRT_31803_slide_13"
Re = "Rebound_33876_slide_36"

 # untreated
Un_df <- read_rds(sprintf("../data/%s_filter.rds", Un))  
Un_df <- Un_df%>%
        mutate(SampleID= sample_Un)%>% # add a column for sampleID
        relocate(SampleID, .after= Image)

# 7days IR
Tr_df <- read_rds(sprintf("../data/%s_filter.rds", Tr))
Tr_df <- Tr_df%>%
        mutate(SampleID= sample_Tr)%>% #add a column for sampleID
        relocate(SampleID, .after= Image)

 # relapse
Re_df <- read_rds(sprintf("../data/%s_filter.rds", Re))
Re_df <- Re_df%>%
        mutate(SampleID= sample_Re)%>% #add a column for sampleID
        relocate(SampleID, .after= Image)

#### All names (columns) in all three groups should have the same format in order to bind them.  

colnames(Re_df) <- gsub("Col.I", "Col I", colnames(Re_df))
colnames(Re_df) <- gsub("VE.Cad", "VE-Cad", colnames(Re_df))
colnames(Re_df) <- gsub("Col.IV", "Col IV", colnames(Re_df))
colnames(Re_df) <- gsub("NF.H", "NF-H", colnames(Re_df))
colnames(Re_df) <- gsub("Lamin.AC", "Lamin AC", colnames(Re_df))
colnames(Re_df) <- gsub ("\\..", ": ", colnames(Re_df))
                   
UnTrRe_df <-rbind(Un_df,Tr_df,Re_df)


# read the metadata which has information about marker used in this experiment
expresion_marker_region <- read_xlsx("../data/HIFI_RT_Marker_Legend.xlsx")

# match the name of the biomarker from the main CSV file and metadata( .xlsx file), REMOVE DAPI, and select only the mean intensities of the biomarker recommended in the metadata ('MFI Measurement Zone')

MarkerRegion <-  unite(expresion_marker_region, col='marker_region', c('Marker', 'MFI Measurement Zone'), sep = " ")
MarkerRegion$marker_region <- gsub(" ", ": ", MarkerRegion$marker_region) 
MarkerRegion$marker_region <- gsub("Nuclei", "Nucleus", MarkerRegion$marker_region)
MarkerRegion$marker_region <- gsub("Col: IV", "Col IV", MarkerRegion$marker_region)
MarkerRegion$marker_region <- gsub("Col: I", "Col I", MarkerRegion$marker_region)




UnTrRe_df <- pivot_longer(UnTrRe_df, cols = "DAPI: Nucleus: Mean": "Phalloidin: Cell: Mean", 
                  names_to = "marker_region", values_to = "Mean_Intensity")


cellposmarkRegion <- left_join(UnTrRe_df ,MarkerRegion, by="marker_region") 

# drop the NA value of the "Channel" column, remove the "DAPI" from the markers
cellposmarkRegion<- cellposmarkRegion %>% 
  drop_na("Channel") %>%
  rename( "Marker" ="Full Name") %>%
  filter(Marker!= "DAPI")  





###################### patient_cell_positions.csv ############
####should read the CSV file which created before for each individual group (the cropped image ) and then combines all group in an object called "CelPos_allgroups"
 
CelPosUn_rectangle <- read.csv(sprintf("../data/patient%s_cell_positions.csv", sample_Un))
CelPosUn_rectangle$SampleID <- sample_Un

CelPosTr_rectangle <- read.csv(sprintf("../data/patient%s_cell_positions.csv", sample_Tr))
CelPosTr_rectangle$SampleID <- sample_Tr

CelPosRe_rectangle <- read.csv(sprintf("../data/patient%s_cell_positions.csv", sample_Re))
CelPosRe_rectangle$SampleID <- sample_Re

CelPos_allgroups <-rbind(CelPosUn_rectangle ,CelPosTr_rectangle,CelPosRe_rectangle)%>%
  select(-c("cluster"))




```




#####  Data preparation and generation of "cellData.csv", which is required for niche-phenotype mapping. in this step, only the functional biomarker which is provided in metadata in a column named `Use For Classification` was selected.

```{r, generation of "cellData.csv", fig.align='center',echo= TRUE}


#Choose the required column from the previously saved data set based on the unique label and filter just the functional biomarker and add a column called SampleID 
cellData <- cellposmarkRegion %>%
  select(label, Marker ,SampleID, Mean_Intensity , `Use For Classification`)%>%
  unique(by=label)%>%
  filter (`Use For Classification` %in% c("No"))

cellData<- cellData %>% 
     pivot_wider(names_from = Marker, values_from =Mean_Intensity )%>%
     select(-c(`Use For Classification`))
    
      

# Rename the column named "label" to "cellLabelInImage" (MIPMAP requirement). Change the biomarker's name to a shorter version. remove unnecessary columns and relocate columns to meet MIPMAP requirements

cellData <- left_join(CelPos_allgroups,cellData, by=c("label","SampleID"))%>%
            rename("cellLabelInImage" ="label", "TNC" = "Tenascin C" , "NF-H"="Neurofilament H",
                   "CC3"="Cleaved caspace 3","WGA"= "Wheat germ agglutinin 647")%>%
            select(-c(x,y,cell_type))%>%
            relocate(SampleID,.before =cellLabelInImage)
       

# Save the data as CSV format with the exact name provided

write_csv(cellData%>%as_tibble(rownames=NA),"../data/cellData.csv" )


```




### load json objects which was saved in "HIFI_UnTrRe_TMENsAnalysis.ipynb" 


```{r, load json objects, echo= TRUE}


DataId = "UnTrRe"
Pc <- fromJSON(file= sprintf("../output/pca_sites_%s.json",DataId))  # PCA on sites 
ArAn <- fromJSON(file= sprintf("../output/ArAn_sites_%s.json" ,DataId)) # Archetype analysis based on sites cell abundance
CeAb <- fromJSON(file= sprintf ("../output/CeAbSi_%s.json", DataId)) # cell abundance of randomly generated sites
CeNi<- fromJSON(file= sprintf("../output/cells_niches_%s.json", DataId)) # sites centered on cells and niches weights
NiCo <-  fromJSON(file= sprintf("../output/niches_color_%s.json", DataId)) #Niches color code 
 

```

#####  NICH_PHENOTYPE_ASSOCIATIONS

```{r, requirement parameter, echo = TRUE }

#Define the requirement parameter for downstream analysis 

CELLTYPES <- c ('Astrocyte', 'Fibroblast', 'Neuron', 'Neutrophil', 'OPC-like','T-cell',
                'TAM', 'Tumor_A','Tumor_B','Vessel','Undefined')

NoNi = 11


```

### Cell abundance in sites

```{r, Cell abundance in sites, fig.align='center',echo= TRUE}

# convert the JSON object to data frame with class tbl_df
SiCeAb <- as_tibble(lapply(CeAb$cellAbSites,unlist))
SiCeAb <- SiCeAb %>%
          select(-c(site_id, patient_id, index))  # remove unnecessary columns



```



##### NICHE IDENTIFICATION 

```{r, NICHE IDENTIFICATION, fig.align='center', echo= TRUE}

# Define an object containing a sequence of "Ni" (niches)and paste  number 1,.. to the length of "NoNi" to "Ni"  
niches <- paste0("Ni",as.vector(seq(1,NoNi,1)))

# Extract the color code which is converted rgb to hex color and saved in "HIFI_NIPMAP.ipynb" as a json file.
NiCo <- as.list(NiCo$Niche_color)

#convert the JSON object to data frame with class tbl_df
NiCePro <- do.call(cbind,lapply(ArAn$nichesCA,unlist))
rownames(NiCePro ) <- CELLTYPES
colnames(NiCePro ) <- niches

# ranspose and pivot the data
NiCePr  <- NiCePro %>%t%>%
  as_tibble(rownames = NA)%>%
  rownames_to_column(var="archetype")%>%
  pivot_longer(cols=all_of(CELLTYPES),names_to="cell_type",values_to = "cell_density")

# Negative cell densities are changed to zero.
NiCePr [NiCePr <0] <-0

```




##### NICHE-PHENOTYPE MAPPING 

```{r, NICHE-PHENOTYPE MAPPING, echo = TRUE }
## Niches weights(proportions) of all cells from all images 

NiInNo <- 2 # number of interface 

MARKERS <-  c("Hif1a", "TNC", "NF-H", "CC3", "Podoplanin", "Laminin", "Periostin", "RedDot2",
             "CSPG5", "Fibronectin", "Col IV", "WGA", "aTubulin", "Col I", "Phalloidin")


CeNi <- as_tibble(lapply(CeNi$cells_niches,unlist))%>%
  rename_at(vars(matches("[0-9]")),~niches)%>%
  mutate(cell_id=as.numeric(cell_id))

# base on the number of given interface and number of niches create the interface of the niches(combination) 

Get_NiIn <- function(IntNo,NiNo){
  NiIn<- combn(niches,IntNo)
  NiIn <- apply(NiIn,2,function(x) paste0(x,collapse=" X "))
  return(NiIn )
}

coreIntf2 <- append(niches,Get_NiIn(NiInNo,NoNi))

```


##### Get markers expression and niche weigths of cells
```{r, markers expression and niche weigths of cells, echo = TRUE }

# df with following columns: cell_type, SampleID, cell_id, Ni1....an & interfaces, marker,value
CePhNi <- read.csv("../data/cellData.csv",check.names=FALSE,header = TRUE, sep =',')%>%
 
  dplyr::rename(cell_id = cellLabelInImage)%>%
  left_join(CeNi%>%filter(cell_type!="Undefined"),.,by=c("SampleID","cell_id"))%>%
  filter(cell_type!="Undefined")%>%
  filter(!(is.na(Ni1)| is.na(Ni2)| is.na(Ni3) | is.na(Ni4)|  is.na(Ni5) |  is.na(Ni6) |  is.na(Ni7) |  is.na(Ni8) |  is.na(Ni9) |  is.na(Ni10) |  is.na(Ni11)))%>%
  mutate("Ni1 X Ni2" = Ni1*Ni2)%>%
  mutate("Ni1 X Ni3" = Ni1*Ni3)%>%
  mutate("Ni1 X Ni4" = Ni1*Ni4)%>%
  mutate("Ni1 X Ni5" = Ni1*Ni5)%>%
  mutate("Ni1 X Ni6" = Ni1*Ni6)%>%
  mutate("Ni1 X Ni7" = Ni1*Ni7)%>%
  mutate("Ni1 X Ni8" = Ni1*Ni8)%>%
  mutate("Ni1 X Ni9" = Ni1*Ni9)%>%
  mutate("Ni1 X Ni10" = Ni1*Ni10)%>%
  mutate("Ni1 X Ni11" = Ni1*Ni11)%>%
  mutate("Ni2 X Ni3" = Ni2*Ni3)%>%
  mutate("Ni2 X Ni4" = Ni2*Ni4)%>%
  mutate("Ni2 X Ni5" = Ni2*Ni5)%>%
  mutate("Ni2 X Ni6" = Ni2*Ni6)%>%
  mutate("Ni2 X Ni7" = Ni2*Ni7)%>%
  mutate("Ni2 X Ni8" = Ni2*Ni8)%>%
  mutate("Ni2 X Ni9" = Ni2*Ni9)%>%
  mutate("Ni2 X Ni10" = Ni2*Ni10)%>%
  mutate("Ni2 X Ni11" = Ni2*Ni11)%>%
  mutate("Ni3 X Ni4" = Ni3*Ni4)%>%
  mutate("Ni3 X Ni5" = Ni3*Ni5)%>%
  mutate("Ni3 X Ni6" = Ni3*Ni6)%>%
  mutate("Ni3 X Ni7" = Ni3*Ni7)%>%
  mutate("Ni3 X Ni8" = Ni3*Ni8)%>%
  mutate("Ni3 X Ni9" = Ni3*Ni8)%>%
  mutate("Ni3 X Ni10" = Ni3*Ni10)%>%
  mutate("Ni3 X Ni11" = Ni3*Ni11)%>%
  mutate("Ni4 X Ni5" = Ni4*Ni5)%>%
  mutate("Ni4 X Ni6" = Ni4*Ni6)%>%
  mutate("Ni4 X Ni7" = Ni4*Ni7)%>%
  mutate("Ni4 X Ni8" = Ni4*Ni8)%>% 
  mutate("Ni4 X Ni9" = Ni4*Ni9)%>%
  mutate("Ni4 X Ni10" = Ni4*Ni10)%>%
  mutate("Ni4 X Ni11" = Ni4*Ni11)%>%
  mutate("Ni5 X Ni6" = Ni5*Ni6)%>%
  mutate("Ni5 X Ni7" = Ni5*Ni7)%>%
  mutate("Ni5 X Ni8" = Ni5*Ni8)%>%
  mutate("Ni5 X Ni9" = Ni5*Ni9)%>%
  mutate("Ni5 X Ni10" = Ni5*Ni10)%>%
  mutate("Ni5 X Ni11" = Ni5*Ni11)%>%
  mutate("Ni6 X Ni7" = Ni6*Ni7)%>%
  mutate("Ni6 X Ni8" = Ni6*Ni8)%>%
  mutate("Ni6 X Ni9" = Ni6*Ni9)%>%
  mutate("Ni6 X Ni10" = Ni6*Ni10)%>%
  mutate("Ni6 X Ni11" = Ni6*Ni11)%>%
  mutate("Ni7 X Ni8" = Ni7*Ni8)%>%
  mutate("Ni7 X Ni9" = Ni7*Ni9)%>%
  mutate("Ni7 X Ni10" = Ni7*Ni10)%>%
  mutate("Ni7 X Ni11" = Ni7*Ni11)%>%
  mutate("Ni8 X Ni9" = Ni8*Ni9)%>%
  mutate("Ni8 X Ni10" = Ni8*Ni10)%>%
  mutate("Ni8 X Ni11" = Ni8*Ni11)%>%
  mutate("Ni9 X Ni10" = Ni9*Ni10)%>%
  mutate("Ni9 X Ni11" = Ni9*Ni11)%>%
  mutate("Ni10 X Ni11" = Ni10*Ni11)%>%
  pivot_longer(cols=all_of(MARKERS),
               names_to="marker",values_to="value")

CM <- correlation_niches_CM(markersCells.niches=CePhNi,Markers=MARKERS,corrMeth="spearman",
                            coreIntf2 ,1/100,0.3,nbNiches=NoNi)

```



##### Plot heatmaps and see the correlation of biomarkers in related cell types, niches, and the interface of niches
```{r,heatmaps, include=FALSE,echo =TRUE }

plot_heatmap_CT(CM.mat=CM, coreIntf2  ,"../output/fig/CMbyCells2.pdf")
plot_heatmap_markers(CM.mat=CM,coreIntf2  ,"../output/fig/CMbyMarkers.pdf")

```



##### TABLE OF NICHE-ASSOCIATED CELL PHENOTYPES 
```{r, TABLE OF NICHE-ASSOCIATED CELL PHENOTYPES , TRUE}
#Archetype weights of randomly generated sites 
ArSi <- as.data.frame(do.call(cbind,lapply(ArAn$alfas,unlist)))

NiNames <- c("Ni1"="Astrocytes","Ni2"="T-cell, TAMs, fibroblasts","Ni3"="T-cell, vessels",
                 "Ni4"="Tumor, low vascularization",
                "Ni5"="OPC-like with a few tumor and neurons",
                "Ni6"="Tumor & astrocyte, high vascularization",
                 "Ni7"="Neuronal","Ni8"="Vessels", "Ni9"= "TAMs with vessel & astrocytes",
                "Ni10"= "Neutrophils, neurons","Ni11"="Empty" )



NiCeAb <- NiCePro %>%t%>%as_tibble(rownames=NA)%>%rownames_to_column(var="archetype")%>%
    mutate(archetype=str_replace_all(archetype,NiNames))

#Since Ni10 and Ni11 are not replaced by the bug in str_replace_all() correctly, we must change the name
NiCeAb$archetype <- gsub("Astrocytes0", "Neutrophils, neurons", NiCeAb$archetype) 
NiCeAb$archetype <- gsub("Astrocytes1", "Empty", NiCeAb$archetype) 

NiCeAb <- NiCeAb %>% column_to_rownames(var="archetype")%>%t()%>%
  as_tibble(rownames=NA)%>%rownames_to_column(var="cell_type")%>%
  filter(cell_type!="Undefined")



###Sort cell types per niche by decreasing cell abundance

NiCeAb.sort <- NiCeAb %>%
   pivot_longer(cols = as.vector(NiNames),names_to="niche",values_to="cell_density")%>%
   group_by(niche)%>%arrange(desc(cell_density))
  



colnames(ArSi) <- niches
ArSiCeAb <- cbind(SiCeAb,ArSi)%>%
  pivot_longer(cols=append(CELLTYPES,"Undefined"),names_to="cell_type",values_to= "cell_density")
 


### Get cell types enriched in each niche (take top 1% sites closest to niche)

NiCeTy <- get_CT_enriched_all_archs(ArSiCeAb ,NiNames ,thresh=0.99)%>%
  group_by(niche)%>%
  mutate(cell_type = paste(unique(cell_type),collapse="\n"))%>%
  distinct(niche,cell_type)


# Since Ni10 and Ni11 are not replaced by the bug in str_replace_all() correctly, we must change the name
NiCeTy$niche <- gsub("Astrocytes0", "Neutrophils, neurons", NiCeTy$niche)
NiCeTy$niche <- gsub("Astrocytes1", "Empty", NiCeTy$niche) 


## Get table of niches/interfaces-associated cell phenotypes

table <-TableNichesPhenotypes(CM=CM,NichesCT=NiCeTy,Niches.names=NiNames ,
                               nichesCA.sorted = NiCeAb.sort,pathFigs = "../output/fig")



#Since Ni10 and Ni11 are not replaced by the bug in str_replace_all() correctly, we must change the name in the table to the desired name and save the table as a pdf.

table$niche <- gsub("Astrocytes0", "Neutrophils, neurons", table$niche)
table$niche <- gsub("Astrocytes1", "Empty", table$niche)



pdf("../output/fig/NichePhenotype_table.pdf",width=20,height=20)
grid.table(table, rows = NULL)

dev.off()
```